---
title: "Plumbing the Data:<br>Mapping U.S. Water Insecurity"
subtitle: "Final project proposal - INFO 526"
author: 
  - name: "Team: Yashi Mi & Nathan Herling"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "**Project description**<br>This project explores county-level water insecurity in the U.S. for 2022 and 2023, focusing on access to complete indoor plumbing. Using data from the American Community Survey (ACS), we analyze geographic and demographic patterns across over 850 counties each year. The goal is to identify regional disparities and year-over-year trends to better understand the social factors linked to plumbing access."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---


## Packages Setup
<details>
  <summary>Installed Packages</summary>
```{r}
#| label: load-pkgs
#| message: false

# - loading package manager
if (!require(pacman))
  install.packages(pacman)

# - loading required packages
pacman::p_load(tidyverse,
               here,
               tibble,
               gt)
```
</details>

## Dataset

```{r}
#| label: Water Insecurity
#| message: false
# Load water insecurity data from the TidyTuesday project (2025-01-28)
# This dataset explores social vulnerability and access to complete indoor
# plumbing across U.S. counties,
# curated by Niha Pereira and featured in the blog post:
# "Mapping water insecurity in R with tidycensus"
# Original data sources include the U.S. Census Bureau (ACS) and the USGS Vizlab’s
# “Unequal Access to Water” visualization.
# Repo: 
# https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-01-28
water_insecurity_2022 <- read_csv(here("water_insecurity",
                                       "water_insecurity_2022.csv"))
water_insecurity_2023 <- read_csv(here("water_insecurity",
                                       "water_insecurity_2023.csv"))
glimpse(water_insecurity_2022)
glimpse(water_insecurity_2023)

```

## About the ACS Plumbing Data
For the final project, we selected the U.S. Water Insecurity dataset from TidyTuesday (2025-01-28). The 2022 and 2023 datasets were collected by the U.S. Census Bureau's American Community Survey (ACS). Each dataset provides comprehensive county-level information, including geographic boundaries, population size, the number of households lacking plumbing, and the percentage of the population lacking plumbing facilities.

The two datastes includes the following variables:

-   `geoid`: The U.S. Census Bureau ACS county id.

-   `name`: The U.S. Census Bureau ACS county name.

-   `year`: The year of U.S. Census Bureau ACS sample.

-   `geometry`: The county geographic boundaries.

-   `total_pop`: The total population.

-   `plumbing`: The total owner occupied households lacking plumbing facilities.

-   `percent_lacking_plumbing`: The percent of population lacking plumbing facilities.

## water_insecurity_2022_data

<details>
  <summary>Code: Water Insecurity 2022 data analysis</summary>
```{r, echo=TRUE, results='hide'}

# Create summary table
table1 <- tibble(
  variable = c(
    "geoid",
    "name",
    "geometry",
    "year",
    "total_pop",
    "plumbing",
    "percent_lacking_plumbing"
  ),
  types = c(
    "character",
    "character",
    "character",
    "numeric",
    "numeric",
    "numeric",
    "numeric"
  ),
  missing_count = c(0, 0, 0, 0, 0, 2, 2),
  missing_percent = c(0, 0, 0, 0, 0, 0.23585, 0.23585)
)
# - format the table with the gt library
gt_table_1 <- table1 |>
  gt() |>
  tab_header(
    #<--- format as markdown
    title = md("Table 1: Variable Summary `water_insecurity_2022_dat`") #<--- format as markdown
  ) |>
  fmt(
    columns = missing_percent,
    fns = function(x) paste0(formatC(x, format = "f", digits = 3), " %")
  ) |>
  cols_label(
    variable = "Variable Name",
    types = "Type",
    missing_count = "Missing Count",
    missing_percent = "Missing (%)"
  ) |>
  tab_style(
    style = cell_text(weight = "bold", align = "center"),
    locations = cells_column_labels(everything())
  ) |>
  tab_style(style = cell_text(align = "center"),
            locations = cells_body(everything())) |> 
  tab_options(
    table.font.size = "small",
    column_labels.font.size = "medium",
    table.width = pct(100)     # <-- full width of container
  )
```
</details>

```{r, echo=FALSE}
# Now render the table by printing the gt object
gt_table_1
```

## water_insecurity_2023_data


<details>
  <summary>Code: Water Insecurity 2023 data analysis</summary>
```{r}
# - manually create a table..
table2 <- tibble(
  variable = c(
    "geoid",
    "name",
    "geometry",
    "year",
    "total_pop",
    "plumbing",
    "percent_lacking_plumbing"
  ),
  types = c(
    "character",
    "character",
    "character",
    "numeric",
    "numeric",
    "numeric",
    "numeric"
  ),
  missing_count = c(0, 0, 0, 0, 0, 1, 1),
  missing_percent = c(0, 0, 0, 0, 0, 0.1171, 0.1171)
)
# - format the table with the gt library
gt_table_2 <- table2 |>
  gt() |>
  tab_header(
    #<--- format as markdown
    title = md("Table 2: Variable Summary `water_insecurity_2023_dat`") 
  ) |>
  fmt(
    columns = missing_percent,
    fns = function(x)
      paste0(formatC(x, format = "f", digits = 3), " %")
  ) |>
  cols_label(
    variable = "Variable Name",
    types = "Type",
    missing_count = "Missing Count",
    missing_percent = "Missing (%)"
  ) |>
  tab_style(
    style = cell_text(weight = "bold", align = "center"),
    locations = cells_column_labels(everything())
  ) |>
  tab_style(style = cell_text(align = "center"),
            locations = cells_body(everything())) |>
  tab_options(
    table.font.size = "small",
    column_labels.font.size = "medium",
    table.width = pct(100)     # <-- full width of container
  )
```
</details>


```{r, echo=FALSE}
# Now render the table by printing the gt object
gt_table_2
```

## Synopsis of data quality:
These two datasets contain data on the lack of complete indoor plumbing in U.S. counties for the years 2022 and 2023. The 2022 dataset includes information from 848 counties, while the 2023 dataset includes 854 counties. Each dataset contains seven variables: geoid, name, geometry, year, total_pop, plumbing, percent_lacking_plumbing. The columns plumbing and percent_lacking_plumbing have a small number of missing values—less than 0.3%—which does not significantly impact the overall analysis.

## Why we chose this data-set:
We chose this dataset because water insecurity is a critical public health issue that may be influenced by various social vulnerability indicators. The data, sourced from the American Community Survey (ACS), provides county-level information on indoor plumbing insecurity for both 2022 and 2023. Since it covers multiple counties and spans two years, it is well-suited for analyzing year-over-year trends and regional disparities across the U.S. Additionally, the dataset is clean, complete, well-documented, and ideal for collaborative analysis and visualization.

## Question:

The two questions to be answered are:

Question 1: Top counties had the highest plumbing insecurity in 2022 and 2023.

Question 2: How did plumbing insecurity change between 2022 and 2023 for each county? How if at all did the change correlate to population change, geographical location, or total owner occupied households lacking plumbing facilities (as represented by the `plumbing` category).

## Analysis plan

The following are the approaches we will be using for each question.

## **Approach for question 1**

-   We want to visualize which U.S. counties experienced the highest level of plumbing insecurity in 2022 and 2023. To represent the available data best, we will be creating bar graphs showing the top 10 counties with the highest percentages of households lacking complete indoor plumbing for each year. The comparison of the plot will be done for each individual year to highlight the most affected regions in the U.S.

    We will clean the data using dplyr package by filtering and selecting the required variables for the analysis, including name, percent_lacking_plumbing, and year. We will sort the counties in descending order of percent_lacking_plumbing for each year. And the Top 10 counties for both 2022 and 2023 will be selected and combined into a single dataset for the visualization.

    To visualize the results, we will use geom_col() from ggplot2 package to create bar graph, showing comparison between 2022 and 2023 using facet separation. These plots will allow us to address our first research question by identifying the top 10 counties with plumbing insecurity in the U.S and comparing the top-affected regions between 2022 and 2023.
    
    
## **Approach for question 2**
- what do you want to visualize?
- clean data, wrangle data, mutate data
- how do you intend to visualize?

::: {.callout-note}
**Note:** Current Proposal 1.0.0. Subject to change.<br>
(version.feature.patch notation)
:::

```{css echo=FALSE}
.callout-note {
  background-color: #d4f0dc; /* light emerald green */
  border-left: 5px solid #2ecc71; /* darker emerald green accent */
  padding: 1em 1.2em;
  border-radius: 5px;
}
```
