---
title: "Proposal title"
subtitle: "Proposal"
author: 
  - name: "Team: Yashi Mi & Nathan Herling"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Project description"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
library(here)
```

## Dataset

```{r}
#| label: Water Insecurity
#| message: false
water_insecurity_2022 <- read_csv(here("water_insecurity","water_insecurity_2022.csv"))
water_insecurity_2023 <- read_csv(here("water_insecurity","water_insecurity_2023.csv"))
glimpse(water_insecurity_2022)
glimpse(water_insecurity_2023)

```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

## Questions

The two questions you want to answer.

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
